---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# chemunits

<!-- badges: start -->
[![R-CMD-check](https://github.com/KopfLab/chemunits/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/KopfLab/chemunits/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/KopfLab/chemunits/graph/badge.svg)](https://app.codecov.io/gh/KopfLab/chemunits)
[![Documentation](https://img.shields.io/badge/docs-online-green.svg)](https://chemunits.kopflab.org/)
<!-- badges: end -->

## About

The **chemunits** package extends the functionality of the **[units](https://r-quantities.github.io/units/)** package, which makes the incredibly powerful [udunits-2](https://github.com/Unidata/UDUNITS-2) API available in R, by implementing two additional features:

 - automatic conversion of units generated from arithmetic (multiplication, division, etc.) to your preferred units
 - automatic scaling of numeric values with units to their ideal SI prefixes (nano-, milli-, kilo-, etc.)
 
This is helpful in all kinds of situations involving units of physical quantities and is particularly powerful in routine calculations for chemical/biochemical/geochemical applications (hence the name "chemunits"). In addition to these features, chemtools also defines molarity (`M = mol/L`) which is not part of the default [udunits-2](https://github.com/Unidata/UDUNITS-2) database given its widespread use in chemical labs and calculations.

Check out the examples below and explore the use of chemunits further in [this vignette](https://chemunits.kopflab.org/articles/explore.html).

## Installation

You can install the development version of chemunits from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("KopfLab/chemunits")
```

## Show me some code

```{r, message = FALSE}
library(chemunits)

# let chemunits know which units you prefer for different types of quantities
# (you can always reuse the same list, just set it once at the beginning of a session)
my_units <- c(
  "m", "g", "mol", "g/mol", # length, mass, amount, molecular weight
  "L", "M", "bar", "M/bar", # volume, molarity, pressure, solubility
  "J", "J/mol" # energy, gibbs free energy
)
chemunits_options(preferred_units = my_units, auto_scale_units = my_units)

# create a tibble and run some fun calculations
library(dplyr)
df <- 
  # define 2 compounds
  tibble(
    compound = c("NaOH", "theobromine"),
    mw = set_cu(c(40, 180), "g/mol")
  ) |>
  # set a set of weights for each
  cross_join(
    tibble(weight = set_cu(c(0.01, 100, 10000), "mg"))
  ) |>
  # calculate the amount and concentration if dissolved
  mutate(
    amount = weight / mw,
    vol = set_cu(200, "mL"),
    conc = amount / vol,
    bla = as.numeric(amount)
  )

# see what this looks like in different print formats
options(digits = 3)
df # default tibble output

# nicely formatted wih units with the values
df |> knitr::kable()
# with units in the header
df |> make_units_explicit() |> knitr::kable() 
```
